Towards Practical First-Order Model Counting

### 2025-01-21 doc/paper/paper.tex
Clean up Section 3.1, adding another termination proof

\begin{example}
  \textsc{Crane} compiles \cref{eq:bijections} for bijection counting into
  \begin{align*}
    \mathcal{E} &= \left\{\,\begin{aligned}f(m, n) &= \sum_{l=0}^{n} \binom{n}{l}{(-1)}^{n-l}g(l, m),\\ g(l, m) &= \sum_{k=0}^{m}[0 \le k \le 1]\binom{m}{k}g(l-1, m-k)\end{aligned}\,\right\};\\
    \mathcal{D} &= \{\, (f, 1) \mapsto \Gamma, (f, 2) \mapsto \Delta, (g, 1) \mapsto \Delta^{\top}, (g, 2) \mapsto \Gamma \,\},
  \end{align*}
  where $\Delta^{\top}$ is a newly introduced domain. (We omit the definition of
  $\mathcal{F}$ as the formulas can get quite verbose.) To compute the number of
  bijections between two sets of cardinality 3, one would evaluate $f(3, 3)$,
  however, the definition of $g$ is incomplete: $g$ is a recursive function
  presented without any base cases. $\mathcal{D}$ encodes that in $f(m, n)$, $m$
  and $n$ represent $|\Gamma|$ and $|\Delta|$, respectively. Similarly, in
  $g(l, m)$, $l$ represents $|\Delta^{\top}|$, and $m$ represents $|\Gamma|$.
\end{example}

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item More emphasis on the motivation for each part of \Cranetwo{} (e.g.,
          how \FindBaseCases addresses key issues in \textsc{Crane}) would add
          depth to the contribution
    \item this section should be expanded by 1 page (``a clear discussion of the
          innovation aspect and technical parts'')
  \end{itemize}
}

@@

\begin{algorithm}[t]
  \caption{\protect\CompileWithBaseCases{$\phi$}}\label{alg:compilewithbasecases}
  \KwIn{formula $\phi$}
  \KwOut{set $\mathcal{E}$ of equations}
  $(\mathcal{E}, \mathcal{F}, \mathcal{D}) \gets \Compile{$\phi$}$\;
  $\mathcal{E} \gets \Simplify{$\mathcal{E}$}$\;\label{line:second}
  \ForEach{base case $f(\mathbf{x}) \in \FindBaseCases{$\mathcal{E}$}$}{
    $\psi \gets \mathcal{F}(f)$\;
    \lForEach{index $i$ such that $x_{i} \in \mathbb{N}_{0}$}{$\psi \gets \Propagate{$\psi$, $\mathcal{D}(f, i)$, $x_i$}$}
    $\mathcal{E} \gets \mathcal{E} \cup \CompileWithBaseCases{$\psi$}$\;\label{line:final}
  }
\end{algorithm}

\todo[inline,caption={}]{
  \begin{itemize}
    \item Reorganise this section (e.g., where do I put the description of
          Simplify?), bringing some TODOs from above.
    \item The algorithm’s recursive nature and its interaction with
          \textsc{Crane} could be explained more clearly
  \end{itemize}
}

\Cref{alg:compilewithbasecases} presents our overall approach for compiling a
formula into equations that include the necessary base cases. To begin, we use
\textsc{Crane} to compile the formula into the three components: $\mathcal{E}$,
$\mathcal{F}$, and $\mathcal{D}$ (as described in \cref{sec:compilation}). After
some algebraic simplification, $\mathcal{E}$ is passed to the \FindBaseCases
procedure (see \cref{sec:identifying}). For each base case $f(\mathbf{x})$, we
retrieve the logical formula $\mathcal{F}(f)$ associated with the function name
$f$ and simplify it using the \Propagate procedure (explained in detail in
\cref{sec:propagating}). We do this by iterating over all indices of
$\mathbf{x}$, where $x_{i}$ is a constant, and using \Propagate to simplify
$\psi$ by assuming that domain $\mathcal{D}(f, i)$ has size $x_{i}$. Finally, on
\autoref{line:final}, \CompileWithBaseCases recurses on these simplified
formulas and adds the resulting base case equations to $\mathcal{E}$.
\Cref{example:bijections} below provides more detail.

\begin{remark*}
  Although \CompileWithBaseCases starts with a call to \textsc{Crane}, the
  proposed algorithm is not just a post-processing step for FOKC because
  \cref{alg:compilewithbasecases} is recursive and can issue more calls to
  \textsc{Crane} on various derived formulas.
\end{remark*}

@@

\begin{example}\label{example:bijectionstwo}
  Let us return to the bijection-counting problem from \cref{example:bijections}
  and its initial solution described in \cref{example:bijectionstwo}. On
  \autoref{line:second}, \Simplify transforms
  $g(l, m) = \sum_{k=0}^{m}[0 \le k \le 1]\binom{m}{k}g(l-1, m-k)$ into
  $g(l, m) = g(l-1, m) + mg(l-1, m-1)$. Then \FindBaseCases identifies two base
  cases: $g(0, m)$ and $g(l, 0)$. In both cases, \CompileWithBaseCases recurses
  on the formula $\mathcal{F}(g)$ simplified by assuming that one of the domains
  is empty. In the first case, we recurse on the formula
  $\forall x \in \Gamma\text{. }S(x) \lor \neg S(x)$, where $S$ is a predicate
  introduced by preprocessing with weights $w^{+}(S) = 1$ and $w^{-}(S) = -1$.
  Hence, we obtain the base case $g(0, m) = 0^{m}$. In the case of $g(l, 0)$,
  \Propagate{$\psi$, $\Gamma$, $0$} returns an empty formula, resulting in
  $g(l, 0) = 1$.

  While these base cases overlap when $l = m = 0$, they remain consistent since
  $0^{0} = 1$. Generally, let $\phi$ be a formula with two domains $\Gamma$ and
  $\Delta$, and let $n, m \in \mathbb{N}_{0}$. Then the FOMC of
  \Propagate{$\phi$, $\Delta$, $n$} assuming $|\Gamma| = m$ is the same as the
  FOMC of \Propagate{$\phi$, $\Gamma$, $m$} assuming $|\Delta| = n$.
\end{example}

@@

\begin{example}
  Consider the recursive function $g$ from \cref{example:bijectionstwo}.
  \FindBaseCases iterates over two function calls: $g(l-1, m)$ and
  $g(l-1, m-1)$. The former produces the base case $g(0, m)$, while the latter
  produces both $g(0, m)$ and $g(l, 0)$.
\end{example}

@@

\begin{theorem}[Termination]\label{thm:halting}
  Let $f$ be an $n$-ary function in $\mathcal{E}$ and
  $\mathbf{x} \in \mathbb{N}_{0}^{n}$. Then the evaluation of $f(\mathbf{x})$
  terminates.
\end{theorem}

@@

\begin{algorithm}[t]
  \caption{A sketch of the C++ program for the equations in \cref{example:bijectionstwo}, particularly highlighting the recursive definition of function $g$.}\label{alg:cpp}
  initialise $\Cache_{g(0, m)}$, $\Cache_{g(l, 0)}$, $\Cache_{g}$, and $\Cache_{f}$\;
  \DontPrintSemicolon
  \textbf{Function} $g_{0,m}(m)$\textbf{:} \dots\;
  \textbf{Function} $g_{l,0}(l)$\textbf{:} \dots\;
  \PrintSemicolon
  \Fn{$g(l, m)$}{
    \lIf{$(l, m) \in \Cache_{g}$}{\Return{$\Cache_{g}(l, m)$}}\label{line:cache}
    \lIf{$l = 0$}{\Return{$g_{0,m}(m)$}}\label{line:basecaseone}
    \lIf{$m = 0$}{\Return{$g_{l,0}(l)$}}\label{line:basecasetwo}
    $r \gets g(l-1, m) + mg(l-1,m-1)$\;
    $\Cache_{g}(l, m) \gets r$\;
    \Return{$r$}\;
  }
  \DontPrintSemicolon
  \textbf{Function} $f(m, n)$\textbf{:} \dots\;
  \PrintSemicolon
  \Fn{\Main}{
    $(m, n) \gets \Input{}$\;
    \Return{$f(m, n)$}\;
  }
\end{algorithm}

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item Experiments could be expanded by 1 page (``a more thorough and
          independent practical assessment'')
  \end{itemize}
}

>>>>

\begin{example}\label{example:solution}
  \textsc{Crane} compiles \cref{eq:bijections} for bijection counting into
  \begin{align*}
    \mathcal{E} &= \left\{\,\begin{aligned}f(m, n) &= \sum_{l=0}^{n} \binom{n}{l}{(-1)}^{n-l}g(l, m),\\ g(l, m) &= \sum_{k=0}^{m}[0 \le k \le 1]\binom{m}{k}g(l-1, m-k)\end{aligned}\,\right\};\\
    \mathcal{D} &= \{\, (f, 1) \mapsto \Gamma, (f, 2) \mapsto \Delta, (g, 1) \mapsto \Delta^{\top}, (g, 2) \mapsto \Gamma \,\},
  \end{align*}
  where $\Delta^{\top}$ is a newly introduced domain. (We omit the definition of
  $\mathcal{F}$ as the formulas can get quite verbose.) To compute the number of
  bijections between two sets of cardinality 3, one would evaluate $f(3, 3)$,
  however, the definition of $g$ is incomplete: $g$ is a recursive function
  presented without any base cases. $\mathcal{D}$ encodes that in $f(m, n)$, $m$
  and $n$ represent $|\Gamma|$ and $|\Delta|$, respectively. Similarly, in
  $g(l, m)$, $l$ represents $|\Delta^{\top}|$, and $m$ represents $|\Gamma|$.
\end{example}

@@

\todo[inline,caption={}]{Add a discussion of the innovation aspect and technical
  parts. Motivate each part of \Cranetwo{} (e.g., how \FindBaseCases addresses
  key issues in \textsc{Crane}).}

@@

\begin{algorithm}[t]
  \caption{\protect\CompileWithBaseCases{$\phi$}}\label{alg:compilewithbasecases}
  \KwIn{formula $\phi$}
  \KwOut{set $\mathcal{E}$ of equations}
  $(\mathcal{E}, \mathcal{F}, \mathcal{D}) \gets \Compile{$\phi$}$\;
  $\mathcal{E} \gets \Simplify{$\mathcal{E}$}$\;\label{line:second}
  \ForEach{base case $f(\mathbf{x}) \in \FindBaseCases{$\mathcal{E}$}$}{\label{line:loopy}
    $\psi \gets \mathcal{F}(f)$\;
    \lForEach{index $i$ such that $x_{i} \in \mathbb{N}_{0}$}{$\psi \gets \Propagate{$\psi$, $\mathcal{D}(f, i)$, $x_i$}$}\label{line:propagate}
    $\mathcal{E} \gets \mathcal{E} \cup \CompileWithBaseCases{$\psi$}$\;\label{line:final}
  }
\end{algorithm}

\Cref{alg:compilewithbasecases} presents our overall approach for compiling a
formula into equations that include the necessary base cases. To begin, we use
\textsc{Crane} to compile the formula into the three components: $\mathcal{E}$,
$\mathcal{F}$, and $\mathcal{D}$ (as described in \cref{sec:compilation}). After
some algebraic simplifications (described below), $\mathcal{E}$ is passed to the
\FindBaseCases procedure (see \cref{sec:identifying}). For each base case
$f(\mathbf{x})$, we retrieve the logical formula $\mathcal{F}(f)$ associated
with the function name $f$ and simplify it using the \Propagate procedure
(explained in detail in \cref{sec:propagating}). We do this by iterating over
all indices of $\mathbf{x}$, where $x_{i}$ is a constant, and using \Propagate
to simplify $\psi$ by assuming that domain $\mathcal{D}(f, i)$ has size $x_{i}$.
Finally, on \autoref{line:final}, \CompileWithBaseCases recurses on these
simplified formulas and adds the resulting base case equations to $\mathcal{E}$.

@@

\begin{example}\label{example:bijectionstwo}
  Let us return to the bijection-counting problem from \cref{example:bijections}
  and its initial solution described in \cref{example:solution}. \Simplify
  transforms $g(l, m) = \sum_{k=0}^{m}[0 \le k \le 1]\binom{m}{k}g(l-1, m-k)$
  into $g(l, m) = g(l-1, m) + mg(l-1, m-1)$. Then \FindBaseCases identifies two
  base cases: $g(0, m)$ and $g(l, 0)$. In both cases, \CompileWithBaseCases
  recurses on the formula $\mathcal{F}(g)$ simplified by assuming that one of
  the domains is empty. In the first case, we recurse on the formula
  $\forall x \in \Gamma\text{. }S(x) \lor \neg S(x)$, where $S$ is a predicate
  introduced by preprocessing with weights $w^{+}(S) = 1$ and $w^{-}(S) = -1$.
  Hence, we obtain the base case $g(0, m) = 0^{m}$. In the case of $g(l, 0)$,
  \Propagate{$\psi$, $\Gamma$, $0$} returns an empty formula, resulting in
  $g(l, 0) = 1$. While these base cases overlap when $l = m = 0$, they remain
  consistent since $0^{0} = 1$.
\end{example}

% Generally, let $\phi$ be a formula with two domains $\Gamma$ and $\Delta$, and
% let $n, m \in \mathbb{N}_{0}$. Then the FOMC of \Propagate{$\phi$, $\Delta$,
%   $n$} assuming $|\Gamma| = m$ is the same as the FOMC of \Propagate{$\phi$,
%   $\Gamma$, $m$} assuming $|\Delta| = n$.

% Although \CompileWithBaseCases starts with a call to \textsc{Crane}, the
% proposed algorithm is not just a post-processing step for FOKC because
% \cref{alg:compilewithbasecases} is recursive and can issue more calls to
% \textsc{Crane} on various derived formulas.

We end this section by proving that \CompileWithBaseCases terminates since each
recursive call on \autoref{line:final} reduces the number of domains in the
formula.

\begin{theorem}
  Given any \FO{} formula $\phi$, \CompileWithBaseCases{$\phi$} terminates.
\end{theorem}

To prove the theorem, we rely on two observations about the algorithms presented
in \cref{sec:identifying,sec:propagating}.

\begin{observation}\label{observation1}
  Each base case returned by \FindBaseCases has at least one constant (in line
  with \cref{def:basecase}).
\end{observation}

\begin{observation}\label{observation2}
  For any formula $\phi$, domain $\Delta$, and $n \in \mathbb{N}_{0}$,
  \Propagate{$\phi$, $\Delta$, $n$} returns a formula with no variables
  quantified over $\Delta$.
\end{observation}

\begin{proof}
  We proceed by induction on the number of domains that variables in $\phi$ are
  quantified over. If there are no domains, then $\phi$ is essentially a
  propositional formula, and \textsc{Crane} compiles it into an equation of the
  form $f = \expr$ with no `function calls'. Suppose that \CompileWithBaseCases
  terminates for all formulas with at most $n \in \mathbb{N}_{0}$ domains. Let
  $\phi$ be a formula with $n+1$ domains. By \cref{observation1}, each base case
  on \autoref{line:loopy} of \cref{alg:compilewithbasecases} has at least one
  constant. Therefore, by \cref{observation2}, after \autoref{line:propagate},
  formula $\psi$ has at most $n$ domains. Thus, \autoref{line:final} terminates
  by the inductive hypothesis, completing the proof that \CompileWithBaseCases
  terminates for an arbitrary formula with $n+1$ domains.
\end{proof}

@@

\begin{example}
  Consider the recursive function $g$ from \cref{example:solution}.
  \FindBaseCases iterates over two function calls: $g(l-1, m)$ and
  $g(l-1, m-1)$. The former produces the base case $g(0, m)$, while the latter
  produces both $g(0, m)$ and $g(l, 0)$.
\end{example}

@@

\begin{theorem}\label{thm:halting}
  Let $f$ be an $n$-ary function in $\mathcal{E}$ and
  $\mathbf{x} \in \mathbb{N}_{0}^{n}$. Then the evaluation of $f(\mathbf{x})$
  terminates.
\end{theorem}

@@

\begin{algorithm}[t]
  \caption{A sketch of the C++ program for the equations in \cref{example:solution}, particularly highlighting the recursive definition of function $g$.}\label{alg:cpp}
  initialise $\Cache_{g(0, m)}$, $\Cache_{g(l, 0)}$, $\Cache_{g}$, and
  $\Cache_{f}$\; \DontPrintSemicolon \textbf{Function} $g_{0,m}(m)$\textbf{:}
  \dots\; \textbf{Function} $g_{l,0}(l)$\textbf{:} \dots\; \PrintSemicolon
  \Fn{$g(l, m)$}{
    \lIf{$(l, m) \in \Cache_{g}$}{\Return{$\Cache_{g}(l, m)$}}\label{line:cache}
    \lIf{$l = 0$}{\Return{$g_{0,m}(m)$}}\label{line:basecaseone}
    \lIf{$m = 0$}{\Return{$g_{l,0}(l)$}}\label{line:basecasetwo}
    $r \gets g(l-1, m) + mg(l-1,m-1)$\; $\Cache_{g}(l, m) \gets r$\;
    \Return{$r$}\; } \DontPrintSemicolon \textbf{Function} $f(m, n)$\textbf{:}
  \dots\; \PrintSemicolon \Fn{\Main}{ $(m, n) \gets \Input{}$\;
    \Return{$f(m, n)$}\; }
\end{algorithm}

@@

\todo[inline]{Expand the section, adding a more thorough and independent
  practical assessment}
